---
title: "Using the TransitionModel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the TransitionModel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE
)
library(IMPACTaf)
```

## Overview

The `TransitionModel` class in the `IMPACTaf_model_pkg` simulates individual patient-level transitions across health states defined by multimorbidity clusters and adverse events. It supports initialization from prevalence and regression-based transitions using a Markov framework.

This vignette shows how to:

- Initialize the model
- Load a Monte Carlo (MC) cohort
- Assign initial states
- Run full transition simulations

---

## Initialize the Model

This loads the YAML configuration, sets up the initial transition matrix, counts MC iterations, and preloads transition probabilities.

```{r eval=FALSE}
tm <- TransitionModel$new()
```

---

## Load and Inspect Cohort

This loads one cohort from a specified Monte Carlo iteration and filters patients aged 65+.

```{r eval=FALSE}
cohort <- tm$get_cohort(mc_counter = 1)
head(cohort)
```

---

## Initialize Patient States

Assigns each patient a starting health state (`CurrentEvent`) based on the initial transition matrix.

```{r eval=FALSE}
lifecourse <- tm$initialize_states(mc_counter = 1)
head(lifecourse)
```

---

## Run the Full Simulation

Runs the Markov model over time and updates each patient's state year by year using transition probabilities.

```{r eval=FALSE}
tm$run_simulation()
```

---

## Output Location

Simulation results are saved as `.csv.gz` files:

```
./outputs/lifecourse/transitions/{mc}_lifecourse.csv.gz
```

Each file contains:
- `pid`, `year`, `CurrentEvent`, `mm_cluster`, `agegrp`, etc.

These can be used to summarize outcomes, visualize disease progression, or feed into cost-effectiveness models.

---

## Output Preview

If desired, you can preview the first few transitions like this:

```{r eval=FALSE}
output <- data.table::fread("./outputs/lifecourse/transitions/1_lifecourse.csv.gz")
head(output)
```

---

## Summary

You’ve just:

- Set up the model
- Loaded patient data
- Assigned initial health states
- Simulated annual transitions over time

For deeper technical reference, see the `?TransitionModel` help page or dive into the source code.



## Running Scenario-Based Simulations

The `TransitionModel` also supports simulation of **alternative scenarios**, where transition probabilities to specific health states can be modified. This is useful for exploring **policy interventions, risk reduction strategies**, or **counterfactual experiments**.

### Defining a Scenario

A scenario is defined by a set of **transition probability adjustments**, passed as a named list to the `run_scenario()` method. For example:

```r
# Create a model object
tm <- TransitionModel$new()

# Define a scenario where the probability of AIS is reduced by 10%
scenario_1 <- list(AIS = 0.9)  # Reduce AIS probability by 10%

# Run the scenario
tm$run_scenario(prob_changes = scenario_1, scenario_name = "reduced_AIS_10pct")
```

Here, the transition probability to `AIS` is multiplied by 0.9 for every row in the transition matrix. The remaining probability mass is redistributed proportionally across all other health states, ensuring that each row still sums to 1.

### How `run_scenario()` Works

For each Monte Carlo iteration:
1. **Loads the basecase transition probability file**
2. **Applies the specified multipliers** to the transition probabilities
3. **Re-normalizes** each row to sum to 1.0
4. **Initializes the cohort** and simulates transitions over time
5. Saves:
   - Adjusted transition probabilities to:  
     `./outputs/transitions/scenarios/{scenario}_{mc}_simulated_probs.csv`
   - Lifecourse data to:  
     `./outputs/lifecourse/transitions/scenarios/{scenario}_{mc}_lifecourse.csv.gz`

### Running Multiple Scenarios

You can define and run multiple scenarios to compare outcomes. For example:

```r
# Scenario 2: Reduce AIS and increase AB probability
scenario_2 <- list(AIS = 0.8, AB = 1.2)

tm$run_scenario(prob_changes = scenario_2, scenario_name = "ais_down_ab_up")
```

### 📈 Analyzing Scenario Outputs

After running a scenario, you can use the output files to:
- Compare prevalence or incidence of events across scenarios
- Assess shifts in health state distributions over time

You can load a scenario’s lifecourse output like this:

```r
library(data.table)

lifecourse_scenario <- fread("./outputs/lifecourse/transitions/scenarios/1_reduced_AIS_10pct_lifecourse.csv.gz")

# Example: Proportion of patients in AIS by year
ais_prev <- lifecourse_scenario[, .(AIS_prev = mean(CurrentEvent == "AIS")), by = year]

ggplot(ais_prev, aes(x = year, y = AIS_prev)) +
  geom_line(color = "firebrick", size = 1.2) +
  labs(title = "AIS Prevalence Over Time (Scenario)", y = "Proportion", x = "Year") +
  theme_minimal()
```

### 🧾 Summary

The scenario analysis framework allows you to:

- Simulate and compare alternative scenarios
- Evaluate robustness of outcomes under varying assumptions

